---
learnr::run_tu---
title: "analysis"
format: html
---

```{r}
library(tidyverse)
library(labelled)
library(haven)
library(knitr)
library(ggplot2)
library(descr)
library(dplyr)
library(summarytools)
library(car)
library(rstatix)
library(psych)
library(lsr)
```

# 💵 Financial Stress in America 💴 {style="color:#FF1F22;"}

**The goal of this project is to answer these questions:**

1\) How does financial stress vary across age groups, particularly among younger adults, in the U.S., and how does it differ by age and level of education?

2\) How has financial stress for different age groups changed from 2007 to 2022?

## View Data

```{r}
# Define the file path (going up one level from /scripts to the main directory)
file_path <- file.path("..", "data", "24data.sav")

# Load the SAV file into a dataframe (automatically a tibble)
df <- read_sav(file_path)

# View the first few rows
head(df)
```

------------------------------------------------------------------------

## 🪙 Individual Frequency Tables 🪙 {style="color:#FF1F22;"}

Here are some of my completed frequency tables: \[Age, Race, and Highest Level of Education, Sex, and Income Level\]

### Demographics: {style="color:#F4AA0B;"}

#### Age (Univariate) {style="color:#C1EF0B;"}

```{r}
#Frequency table for age
library(dplyr)

#Age groups
df <- df %>%
  mutate(
    Age_Grouped = case_when(
      DMAGE >= 18 & DMAGE <= 29 ~ "18-29 (Younger Adults)",
      DMAGE >= 30 & DMAGE <= 49 ~ "30-49 (Middle Adults)",
      DMAGE >= 50 & DMAGE <= 64 ~ "50-64 (Older Adults)",
      DMAGE >= 65              ~ "65+ (Seniors)",
    )
  )

freq(df$Age_Grouped, plot = TRUE)

ggplot(df, aes(x = Age_Grouped)) +
  # Create the bar geometry (using a slightly different color for distinction)
  geom_bar(fill = "#2ECC71", color = "black", alpha = 0.8) +
  
  # Add counts above the bars for clarity
  geom_text(stat='count', aes(label=after_stat(count)), 
            vjust=-0.5, size=3.5) +
  
  # Set up labels and title
  labs(
    title = "Distribution of Survey Respondents by Age Group",
    subtitle = "A critical variable for longitudinal financial stress analysis",
    x = "Age Group",
    y = "Frequency (Number of Respondents)"
  ) +
  
  # Apply a clean, academic theme and adjust text
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 15, hjust = 1) # Angle labels for clarity
  )
```

**Interpretation:** By looking at frequency table of age groups, I can identify how many respondents fall into each age bracket. This allows you to see whether younger adults are heavily represented and to focus analysis on them. I want to figure out if younger adults experience higher financial stress than older populations. From the graph I see that middle aged adults tended to be more reported in the survey.

**Management**: The original variable, `DMAGE`, was a continuous variable but I wanted to analyze age groups instead of individual years, so I used the `mutate()` and **`case_when()`** functions from the `dplyr` package. This created `Age_Grouped`, which binned the respondents into four distinct groups (18-29, 30-49, 50-64, and 65+). The new variable is necessary for specifically examining financial stress among younger adults.

**Graphing**: Since this is a categorical variable. I choose to present this with a bar chart.

------------------------------------------------------------------------

------------------------------------------------------------------------

#### Highest Level of Education (Univariate) {style="color:#C1EF0B;"}

```{r}
#Frequency table for Highest Level of Education
df <- df %>%
  mutate(
    Education_Condensed = case_when(
      DMEDUUS %in% 1:3 ~ "High School Diploma or Less", 
      DMEDUUS %in% 4:5 ~ "Some College or Training",
      DMEDUUS %in% 6:7 ~ "Associate or Bachelor's Degree", 
      DMEDUUS %in% 8:9 ~ "Graduate Degree",
      TRUE ~ NA_character_ 
    ),
    
    # ordered factor for plotting
    Education_Condensed = factor(Education_Condensed,
                                 levels = c("High School Diploma or Less", 
                                            "Some College or Training",
                                            "Associate or Bachelor's Degree", 
                                            "Graduate Degree"),
                                 ordered = TRUE)
  )

ggplot(df, aes(x = Education_Condensed)) +
  # Create the bar geometry
  geom_bar(fill = "#5DADE2", color = "black", alpha = 0.8) +
  
  # Add counts above the bars for clarity
  geom_text(stat='count', aes(label=after_stat(count)), 
            vjust=-0.5, size=3.5) +
  
  # Set up labels and title
  labs(
    title = "Distribution of Respondents by Condensed Education Level",
    subtitle = "Grouping DMEDUUS into Four Attainment Categories",
    x = "Education Level",
    y = "Frequency (Number of Respondents)"
  ) +
  
  # Apply a clean, academic theme and adjust text for readability
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 15, hjust = 1) # Slight angle for long labels
  )
```

**Interpretation:** Education level often correlates with financial security and earning potential. Examining the distribution of education categories helps me evaluate whether stress is concentrated among those with lower educational attainment or if higher education buffers financial stress in modern America.

**Management**: Similarly, the `DMEDUUS` variable contained numeric codes for education level. I used the **`factor()`** function again to re-label these codes (1 through 9) into their corresponding descriptive categories.

------------------------------------------------------------------------

### Financial Stress: {style="color:#F4AA0B;"}

#### Average Money Stress (Univariate) {style="color:#C1EF0B;"}

```{r colnames}
df <- df %>%
  mutate(
    #ordered factor (1 to 10 scale)
    Stress_Score = factor(Q605, 
                          levels = 1:10, 
                          ordered = TRUE)
  )

# --- 2. Visualization (Bar Chart/Frequency Plot) ---
ggplot(df, aes(x = Stress_Score, fill = Stress_Score)) +
  # Use geom_bar to show frequency for each discrete stress level
  geom_bar(color = "black", show.legend = FALSE, width = 0.8) +
  
  # Add counts above the bars
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-0.5, size=3) +
  
  # Apply a gradient color scale (optional, but visually appealing)
  scale_fill_manual(values = colorRampPalette(c("#C8E6C9", "#E57373"))(10)) +
  
  # Set up labels and title
  labs(
    title = "Distribution of Average Money Stress in the U.S.",
    subtitle = "Reported Stress Levels on a 1 (Low) to 10 (High) Scale",
    x = "Average Money Stress Score (1 = Lowest Stress, 10 = Highest Stress)",
    y = "Frequency (Number of Respondents)"
  ) +
  
  # Apply a clean, academic theme
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.x = element_text(size = 10)
  )
```

------------------------------------------------------------------------

## Univariate Graphs

```{r}
age_data_summary <- df %>%
  # Filter out NA's before counting
  filter(!is.na(Age_Grouped)) %>% 
  group_by(Age_Grouped) %>%
  summarize(Count = n()) %>%
  ungroup()

# --- 3. Visualization (Point and Line Plot) ---
ggplot(age_data_summary, aes(x = Age_Grouped, y = Count, group = 1)) +
  
  # Add the line component
  geom_line(color = "#3498DB", linewidth = 1.2) +
  
  # Add the point component (to mark the exact count for each group)
  geom_point(color = "#2980B9", size = 4) +
  
  # Add labels above the points
  geom_text(aes(label = Count), 
            vjust = -1,  # Position the text above the point
            size = 4, 
            color = "#2C3E50") +
  
  # Set up labels and title
  labs(
    title = "Respondent Frequency Across Age Groups",
    subtitle = "A Line Chart Representation of Age Distribution",
    x = "Age Group",
    y = "Number of Respondents (Frequency)"
  ) +
  
  # Apply a clean theme and adjust text
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 15, hjust = 1) # Angle labels for clarity
  )
```

**Interpretation:** By looking at frequency table of age groups, I can identify how many respondents fall into each age bracket. This allows me to see whether younger adults are heavily represented and to focus analysis on them. I want to figure out if younger adults experience higher financial stress than older populations. From the graph I see that middle aged adults tended to make up the largest proportion of the sample while younger adults are the smallest group.

```{r colnames}
library(dplyr)
library(ggplot2)

ggplot(df, aes(x = Age_Grouped, y = Q605, fill = Age_Grouped)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  labs(
    title = "Average Level of Stress by Age Group",
    x = "Age Group",
    y = "Level of Stress"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")
```

**Interpretation:** My explanatory variable is the age group, my response variable is the level of housing cost stress. As age increases, average housing cost stress tends to decrease, suggesting older adults may experience less housing-related financial strain than younger adults.

# Hypothesis Testing

Question: How is financial stress related to educational attainment in America, and are individuals with less formal education more likely to experience financial strain?

Sample Description: This study will focus on a representative sample of U.S. adults and their reported levels of financial stress. The analysis will consider respondents’ educational attainment and their self-reported stress scores to explore patterns of financial strain across education levels.

**Variables**:

-   Explanatory Variable: Educational Attainment (DMEDUUS \[Level_of_Education\] & Age

    -   Categorical Type

-   Responsive Variable: Financial Stress (Average Stress Rating as of 2023) \[\$Q605\]

    -   Quantitative Type

**Null Hypothesis (H₀):** There is no relationship between educational attainment and financial stress among U.S. adults.

**Alternative Hypothesis (Hₐ):** There is a relationship between educational attainment and financial stress among U.S. adults.

Since educational attainment is a categorical variable, for its multiple education groups, and financial stress is a quantitative variable, the appropriate statistical test is an ANOVA test.

# ANOVA

### Set Up

```{r}
# Define the file path
file_path <- file.path("..", "data", "24data.sav")
df <- read_sav(file_path)

# --- CLEANING & VARIABLE CREATION ---
df <- df %>%
  mutate(
    # 1. Age Grouping
    Age_Grouped = case_when(
      DMAGE >= 18 & DMAGE <= 29 ~ "18-29 (Younger Adults)",
      DMAGE >= 30 & DMAGE <= 49 ~ "30-49 (Middle Adults)",
      DMAGE >= 50 & DMAGE <= 64 ~ "50-64 (Older Adults)",
      DMAGE >= 65              ~ "65+ (Seniors)",
      TRUE ~ NA_character_
    ),
    Age_Grouped = factor(Age_Grouped,
                         levels = c("18-29 (Younger Adults)", "30-49 (Middle Adults)", 
                                    "50-64 (Older Adults)", "65+ (Seniors)"),
                         ordered = TRUE),
    
    # 2. Education Condensation
    Education_Condensed = case_when(
      DMEDUUS %in% 1:3 ~ "High School Diploma or Less", 
      DMEDUUS %in% 4:5 ~ "Some College or Training",
      DMEDUUS %in% 6:7 ~ "Associate or Bachelor's Degree", 
      DMEDUUS %in% 8:9 ~ "Graduate Degree",
      TRUE ~ NA_character_
    ),
    Education_Condensed = factor(Education_Condensed,
                                 levels = c("High School Diploma or Less", "Some College or Training",
                                            "Associate or Bachelor's Degree", "Graduate Degree"),
                                 ordered = TRUE),
    
    # 3. Financial Stress Score (Dependent Variable)
    # FIX: Robustly convert to numeric, essential for ANOVA calculations
    Stress_Score_Numeric = as.numeric(as.character(Q605)) 
  ) %>%
  # Filter out missing values for the ANOVA
  filter(!is.na(Stress_Score_Numeric) & 
         !is.na(Age_Grouped) & 
         !is.na(Education_Condensed))
```

### Descriptive

```{r}
xtabs(~ Age_Grouped + Education_Condensed, data = df)

# Mean Average Stress Score by Age Group
aggregate(Stress_Score_Numeric ~ Age_Grouped, data = df, mean)

# Mean Average Stress Score by Education Group
aggregate(Stress_Score_Numeric ~ Education_Condensed, data = df, mean)

# Overall descriptive stats for the dependent variable (Average Stress)
psych::describe(df$Stress_Score_Numeric)
```

```{r}
ggplot(df, aes(x = Age_Grouped, y = Stress_Score_Numeric, color = Education_Condensed, group = Education_Condensed)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) + # Standard Error
  labs(
    title = "Mean Financial Stress by Age and Education (Interaction Plot)",
    x = "Age Group",
    y = "Average Money Stress Score (1-10)",
    color = "Education Level"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1),
        plot.title = element_text(face = "bold", hjust = 0.5))

```

## Factorial Anova

#### Model

```{r}
model_anova <- aov(Stress_Score_Numeric ~ Age_Grouped * Education_Condensed, data = df)

```

#### Assumptions

```{r}
df$Group_Combo <- interaction(df$Age_Grouped, df$Education_Condensed, sep = "_")
leveneTest(Stress_Score_Numeric ~ Group_Combo, data = df)
```

```{r}
df$Group_Combo <- interaction(df$Age_Grouped, df$Education_Condensed, sep = "_")
leveneTest(Stress_Score_Numeric ~ Group_Combo, data = df)
```

```{r}
sample_residuals <- sample(residuals(model_anova), 500)
shapiro.test(sample_residuals)
```

#### Model Summary

```{r}
# Main ANOVA Table
summary(model_anova)

# Effect Size (Eta Squared)
# Measures the proportion of total variance explained by each factor
etaSquared(model_anova)
```

#### Post-Hoc Analysis (because we have differences in groups)

```{r}
# Run Tukey's Honest Significant Difference (HSD) test on the model
TukeyHSD(model_anova)
```

## Interpretation:

The two-way Analysis of Variance (ANOVA) revealed that both age group and education level have a significant effect on average money stress, but the interaction between these two factors was not significant.

-   **Interaction Term:** The interaction between Age Group and Education Condensed was **not statistically significant** (F(9, 3155) = 1.61, p = 0.106). This confirms that the effect of age on financial stress is consistent across all levels of education.

-   **Primary Predictor (Age):** There was a highly significant main effect of Age Group on financial stress, F(3, 3155) = 141.07, p \< 0.001. This effect size is considered large, indicating age is the strongest predictor of financial stress in this model.

-   **Secondary Predictor (Education):** There was a statistically significant main effect of Education Condensed, F(3, 3155) = 3.12, p = 0.025. However, the effect size is **very small**, suggesting educational attainment plays a minor role overall.

Detailed Post-Hoc Analysis (Tukey HSD)

Since both main effects were statistically significant, Tukey's Honest Significant Difference (HSD) test was performed to identify which specific group means differed.

1\. Age Group Differences

Financial stress generally decreases as age increases. The Tukey test revealed significant differences between nearly all consecutive age groups (p \< 0.001 for all), confirming a robust age gradient in financial stress:

-   **18-29 (Younger Adults)**: Showed significantly higher stress than the **50-64** and **65+ (Seniors)** groups.

-   **30-49 (Middle Adults)**: Showed significantly higher stress than the **50-64** and **65+ (Seniors)** groups.

-   **50-64 (Older Adults)**: Showed significantly higher stress than the **65+ (Seniors)** group.

-   **Non-Significant Comparison:** The only non-significant difference in stress was between the **18-29** group and the **30-49** group (p=0.678), suggesting these two younger cohorts experience a similarly high level of financial strain.

2\. Education Condensed Differences

Despite the overall significant F test, the Tukey HSD revealed only one specific significant pairwise difference in education (at a=0.05\$):

-   Individuals with **Some College or Training** reported **significantly higher** average money stress than those with an **Associate or Bachelor's Degree** (p=0.041).

-   All other comparisons (e.g., Graduate Degree vs. High School Diploma or Less) were not statistically different.

### Overall Conclusion

The null hypothesis, "There is no relationship between educational attainment and financial stress," is rejected. However, the alternative hypothesis must be refined:

**Final Result:** The analysis strongly supports the conclusion that **age is the primary demographic predictor of financial stress**, with younger and middle-aged adults reporting significantly higher stress than older adults. Educational attainment shows a complex, minor influence, primarily limited to higher stress in the 'Some College' group compared to those who completed an associate/bachelor's degree.
